---
title: "Intro to Plotting ELPA Data"
author: "Albert Crosby"
date: '2022-06-23'
email: acrosby@sdale.org
output:
      rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Intro to Plotting ELPA Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,fig.align="center")
```

```{r load-data, message=FALSE}
library(elpa)
library(dplyr)
library(ggplot2)
library(ggalluvial)
library(rmarkdown)

# Reading the ELPA data for 2020 from a CSV file:

#elpa2020<-read.csv(file="c:/users/acrosby/Documents/ELPA2021/elpaspring2020.csv") %>%
# Pick the needed fields from the CSV and give expected names
  # select(
  #   Student.ID=StudentID,
  #   Enrolled.Grade=EnrolledGrade,
  #   School=EnrolledSchool,
  #   Ethnicity=Ethnicity,
  #   Scale.Score=ELPA21SummativeScaleScore,
  #   Performance.Level=ELPA21SummativePerformanceLevel
  # )%>%
  # mutate(SY='2020')%>%
  # helper.cleandata()

# Read the ELPA 2021 files downloaded and extracted into a specific folder 
# elpa2021<-readelpascores("C:/Users/acrosby/Downloads/ELPA2021/")%>%
#   helper.reducedata(SchoolYear='2021') %>%
#     helper.cleandata()

# Read the ELPA 2022 files downloaded and extracted into a specific folder 
# elpa2022<-readelpascores("C:/Users/acrosby/Downloads/ELPASpring2022/")%>%
#   helper.reducedata(SchoolYear='2021') %>%
#   helper.cleandata()
# 
# elpa.sdale<-bind_rows(elpa2020,elpa2021,elpa2022)

######################################################
# Combine the dataset(s) from above into a data.table
# named elpadata - that could/should be loaded already?
######################################################

```

## NOTE
This guide was not originally intended as a package vignette.  It may have some layout 
and other issues.

IT IS NOT COMPLETE AT THIS TIME AND IS DEFINITELY A WORK IN PROGRESS.  IT IS INCLUDED
PARTIALLY AS A PLACEHOLDER, AND A WAY TO DEMONSTRATE AT LEAST SOME OF THE FUNCTIONS.

The charts do not look at all similar to what they appear in the original document.  If
I can figure out how to simply load the PDF as a vignette with the R package, that's 
what I will do...

## Introduction

This guide is intended as a walk-through for an evaluation of ELPA21 results for multiple years.  The data
is retrieved directly from the ELPA21 portal (or other source that can supply the needed information).  This
document, and the charts and tables it includes, was created using the R programming environment.  This document can
also serve as the vignette (introductory document) for the elpa21 package that will be available for you to install 
and use.

In this guide, we'll be using ELPA21 assessment results for Springdale Public Schools in Springdale, Arkansas
for three administrations of the test:

* Spring 2020, in the 2019-2020 School Year
* Spring 2021, in the 2020-2021 School Year
* Spring 2022, in the 2021-2022 School Year

Sample charts are included and explored, the programs used are introduced, and guidance is provided on
recreating the results with data from other districts or state wide.



### English Learners

Language Minority (LMS) Students are any students who live in a household where languages other than
English is spoken on a regular basis.

English Learners are Language Minority Students who have a documented need for supportive intervention
demonstrated on one or more measurement tools.

The primary measurement tool in Arkansas is the ELPA21 English Language Proficiency Assessment.

In Springdale Public Schools, over 50% of the student body is considered "Language Minority Students", and 
between 30% and 40% of the entire student body is idenitfied as English Learners and receive supportive ELL
intervention services.

Most of Springdale's LMS students (and those receiving language services) are recent immigrants to the 
United States.
They also experience high rates of mobility as families relocate or return to their country of origin.

There are two major groups of LMS students in Springdale:  Hispanic/Latino students who speak Spanish, and
Islander students, primarily from the Marshall Islands who speak Marshallese.

### ELPA21

ELPA21 is an assessment that measures English learner's mastery of communication skills needed to meet
academic standards.  ELPA21 is created and managed by consortium of states, housed at the National Center for 
Research, Evaluation, Standards and Student Testing at the University of California, Los Angeles.
You can learn more about the ELPA21 suite of assessments at <https://elpa21.org>.

Arkansas, like most states using ELPA21,administers two versions of the ELPA21 test.  

The ***Screener*** is given to English 
learners when they arrive in the state for the first time, and is used to determine if they need 
supportive intervention services to ensure their academic success. 

The ***Summative*** ELPA21 is given to every Arkansas student receiving English Language services
in the Spring semester -- usually sometime between early February and mid April.  
The Summative exam assesses 
skills in four domains: listening, reading, speaking and writing.  As well as scores in each domain, 
learners also receive an Overall Proficiency Level and an Overall Comprehension Level.

Students who receive a "Proficient" Overall Proficiency Level are considered for exiting services, along
with other factors such as performance on standardized tests and classroom performance.

The ELPA21 test **does not** measure academic performance.  It measures the ability of an English Learner
to access coursework and achieve academic progress at the same rate as their age and grade level English
fluent peers.

The charts and evaluations in this guide use the Summative test results, specifically the Overall Proficiency Level.

### R

This guide, and the entire project, makes extensive use of the **R** environment.

**R**, one of the world's leading statistical software suites, is available completely for free.  It's also
fairly easy to read, and there are many excellent guides available.

**R** can read and write data from different sources, such as CSV or Excel, calculate simple summaries or 
statistical models to explore the data, and create simple or complex graphics to give insight to
the data.

The **R** environment is more than just a programming language. There are several graphical
editors, a batch mode, hundreds of community contributed "packages" that extend its functionality, and even
a web server to allow interactive data exploration on-line.

**R** is created, "owned" and maintained by a non-profit organization: [The R Foundation](https://r-project.org).
**R** is distributed as Free Software under the terms of the Free Software Foundation's GNU General Public
License.  You can learn more about R on their web page at <https://r-project.org>.

This guide will not attempt to guide you through writing an **R** program -- but will help you *execute* the
instructions needed to recreate these charts with your own data.

The **R** environment is often compared to the **SAS** System, which you may have used in a university
course.  **SAS** can do most of the things that R can; but unlike **R**, **SAS** comes with commercial
support ***and*** a hefty price tag.

### RStudio

RStudio is a popular (and powerful) graphical front end for the R environment.  Much like R itself, 
RStudio is
available completely for free -- though the company behind RStudio also offers paid support and enhanced
products for a fee (that allows them to continue existing).

If want to recreate this
document using your
own data, I strongly recommend starting with installing the free version of RStudio Desktop.

You can download RStudio at <https://www.rstudio.com/products/rstudio/>

RStudio Desktop is available for Windows, MacOS and Linux.

### R Markdown

This document was created with R Markdown. Markdown is a simple formatting syntax for authoring HTML, PDF,
and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

An R Markdown document allows you to create high quality dynamic reports.  You can mix plain text
with R programming code so that one file contains the descriptive text *and* the supporting statistics,
tables and charts.

While this guide itself is written in RMarkdown, and the files used for creating it are available, teaching
RMarkdown itself is a task beyond it's scope.

All of the charts included "in-line" as figures in this document can also be generated as full-size pages
for printing or including in a slide show.

In the section that discusses how to recreate these charts on your own, there will be more detail on how to 
generate the included charts and tables using R and RStudio..

\newpage

## Exploring the ELPA21 Data Visually

Once the data of ELPA21 scores is loaded, a series of simple evaluation charts can be created. The actual 
loading of the data will be covered in a later section.

### Number of Students Tested Each Year

Figure \ref{fig:ntested} simply shows the number of students tested each year, across all schools and 
grades on a "bar
chart".  Each bar is divided into sections showing the Overall ELPA Proficiency Level, and labelled with
the number of students assessed at that level:

```{r ntested,fig.cap="Number of Students By Year"}
elpa.plot.ntested(elpadata,label.size=4)+ggtitle("Number of Students Tested By Year")
```

Of course, that can be broken down in several different ways.  In R, we can "facet" graphs by different
distinctions such as "Grade Band" (Figure \ref{fig:n-band}):

```{r n-band,fig.cap="Number of Students By Year and ELPA Grade Band"}
  elpadata %>% mutate(GradeBand=helper.gradeband(Enrolled.Grade))%>%
    elpa.plot.ntested(label.size=2)+ 
      facet_grid(GradeBand~.,as.table=F,switch='both')+
      theme(axis.text.y=element_text(),axis.line.y=element_blank(),axis.ticks.y=element_blank())+
      ggtitle("ELPA Students Tested By Grade Band")
```

The Grade Bands used in this chart reflect the testing grade bands of the ELPA21 Summative Assessment itself.

\newpage

## Examining the Distribution of ELPA Scores

The students who take the ELPA test each year changes based on who enters the district as an English Learner
and who gains sufficient English language skills to be exited from services.  This
implies that the distribution of scores over time may change due to factors beyond the district's control.

A "histogram" is a kind of bar chart that shows the number of students who scored within a specified range.
Figure \ref{fig:histogram} is a histogram of the scaled score for the students tested in 
Springdale for all three years.

```{r histogram,fig.cap="Distribution of Scaled Scores By Year"}
elpa.plot.histogram(elpadata)
```

A guick review of Figure \ref{fig:histogram} suggests that the distribution of scores in Springdale may be 
shifting to the left, lower skills, over time. This
could be caused by any number of factors, including exiting more highly skilled students from services, and
an influx of new arrivals with limited English skills.

The ELPA21 consortium cautions that it is not appropriate to compare Scale Scores across the Grade Bands of
the test, so Figure \ref{fig:histogram} may lead to some inaccurate conclusions.

R allows us to facet a chart by multiple dimensions.  In Figure \ref{fig:hist-band}, we can see the
distribution of Scaled Score by the ELPA21 Grade Bands.

```{r hist-band,fig.cap="Distribution of Scaled Scores By Year and Grade Band"}
elpadata %>% mutate(GradeBand=helper.gradeband(Enrolled.Grade))%>%
  elpa.plot.histogram(binwidth=100) + facet_grid(GradeBand ~ SY)

```

Figure \ref{fig:hist-band} allows us to compare across years and across grade bands.  There are still some
difficulties in drawing conclusions from this chart:  each faceted plot is rather small, and the number of
students tested in each band varies.  That variation in n size makes it hard to know if a greater 
*proportion* of students are scoring lower or higher.

There is a statistical approach called a "Density Chart" that addresses this issue.  It's similar to a
histogram, but instead of bars for counts, it draws a smoothed curve where all of the area under the
curve adds
to 100%.  Figure \ref{fig:density-all} shows the density distribution for all 3 years by grade level overlaid.

```{r density-all,fig.cap="Scaled Score Density Distribution by Year and Grade"}
 elpa.plot.density.allyears(elpadata)
```

The density distributions in Figure \ref{fig:density-all} are much easier to compare and draw potential
conclusions from than the histograms in Figure \ref{fig:hist-band}.  The distributions, within each grade
level, look substantially the same for most combinations.

A Z-test can be used to determine the similarity of two density distributions. 
Figure \ref{fig:denscomp} shows that comparison for School Years 2021 and 2022, by grade level.

```{r denscomp, message=FALSE,fig.cap="Scaled Score Density Distribution Comparison"}
elpa.plot.density.compare(elpadata,2021,2022)
```

In Figure \ref{fig:denscomp}, we observe that, when comparing 2021 scores to 2022 scores, there was no
significant difference in the distributions for most grades.  The distribution of Kindergartner scores
on the Spring Summative assessment is significantly different, which suggests that further investigation
might be justified.

\newpage
## Comparing Testing Across Years

There are three primary reasons to examine ELPA21 test results: 

1. Determine the language skills of students being taught in the current year
2. Identify changes in the language skills demographic of our school population for planning purposes
3. Reflect on the success or challenge in moving students forward in the path to English proficiency and future academic success

Charts that show the number of students tested each year, like Figures \ref{fig:ntested}, \ref{fig:n-band}
and \ref{fig:n-eth} only truly address the first two reasons, but fail to show the progress made in moving
students, individually or in aggregate, towards English proficiency.

```{r n-eth,message=FALSE,fig.dim=c(7,2.5),fig.cap="Number tested by Year and Ethnicity"}
  elpa.plot.ntested(elpadata)+ggtitle("Number of Students Tested By Year/Race")+
    facet_grid(SY ~ 
                 factor(case_when(Ethnicity=='Hispanic'~ 'Hispanic',Ethnicity=='Islander'~'Islander', TRUE~'Other')),
               switch="both",scales="free",space="fixed")

```

In order to properly reflect on progress moving students to proficiency, it is important to consider 
students who tested in both years, and what change they experienced over that time period.

\newpage

## The ELPA21 "Flow" Chart

```{r FlowChart,message=FALSE}
elpa.plot.flow(elpadata,2021,2022,incN=TRUE,labelflow=TRUE)
```

\newpage

## The ELPA Dataset

The project expects a data set with the the following variables:

- Student.ID\
A linking ID for the student records, across years.  The value is not displayed in any charts.
In Arkansas, the "triand"/"state testing id" for the student is used.  Stored as a character string
- Enrolled.Grade\
The enrolled grade from the ELPA file\
Levels: "KG","01","02","03"..."12"
- School\
The enrolled school from the ELPA file, as a string.  The values should be the same for all years.  A helper function
is included that strips non-alphabetic characters from the string.
- Ethnicity\
The Ethnicity as reported in the ELPA file, as a string.  Values should be the same for all years.  A helper function
is included to convert several common ethnicity descriptions into a standardized form.
- Scale.Score\
The overall summative scaled score, stored as an integer
- Performance.Level\
The overall summative performance level, stored as a string.\
Levels: "Emerging","Progressing","Proficient"
- SY\
The school year, in form "2022"
Levels: c("2020","2021","2022")


